Cover letter for ``Two applications\dots''

### 2021-03-06 papers/types2020/revisions/sec3.tex
Working on the fpc section.

\begin{figure}[t]
\centering
  \[
\infer{\XXi_1\vdash k\ l : A}
      { \mathrm{Ind}[p] \ (\Gamma_I := \Gamma_C) \in E\quad 
       (\mathit{head} \ A) : T \in \Gamma_I\quad  
       k : D \in \Gamma_C\quad 
       \bc{\XXi_2}{} l D A \quad\unfoldExpert{\XXi_1}{\XXi_2}} 
\]
\vskip -18pt
\[
  \infer{\XXi\vdash  A : s}{E[] \vdash_{CIC} A : s\quad \sortExpert{\XXi}}% The order should be s : A, not A : s ?
  %% we are checking nat : Set, for example, so A : s --AM
  \qquad
  \infer{\bc{\XXi}{}{[]} A A'}{E[]\vdash_{CIC} A =_{\beta\delta\iota\xi}A'\quad\initLExpert{\XXi}}
\]
  \vskip -18pt
\[
  \infer{\bc{\XXi}{}{(t\ ::\ l)}{\depprod{x}{B}{D}}{A}}
        {\XXi_1\vdash t : B\quad
         \bc{\XXi_2}{}l {D[t/x]} A\quad
         \prodE{\XXi}{\XXi_1}{\XXi_2}{t}}
\]
\caption{Specification of a core calculus% of the checking of a proof certificate and the
  % synthesis of a dependently typed $\lambda$-term
  .}
\label{fig:augmented}
\end{figure}

@@

The proof system is parameterized by Coq's global environment $E$,
here a set of constant and inductive definitions; following Coq's
reference manual, inductive definition are denoted by
$\mathrm{Ind}[p] (\Gamma_I := \Gamma_C)$, where $\Gamma_I$ determines
the names and types of the (possibly mutually) inductive type and
$\Gamma_C$ the names and types of its constructors; finally $p$
denotes the number of parameters and plays here no role. We delegate to
Coq's type checking the enforcement of the well-sortedness of
inductive types. The \emph{local context} is empty, since we are only
dealing with types that correspond to Horn clauses, and atomic types
are inductively defined. In fact, we do not have a $\forall$ rule on
the right, although the proof theory would gladly allow it. This means
at the term level that there are no variables in our grammar of
terms. Terms are always applied to a (possibly empty) list of
arguments.  The \emph{decide} rule, % is the equivalent in our
% a \emph{decide} rule in an explicitly focused system
% \cite{liang09tcs}.  
as in the previous proof system for Horn logic,  given an atom,
 selects a clause on which to back-chain on: we lookup the
constructors of an inductive definition from the global
environment, one that matches the head symbol of the atom we aim to backchain on, and then
call the latter judgment that will find a correct instantiation, if any. The rules for backchaining include
 the (conflation of the) left introduction rules for
$\forall$ and $\supset$, as well as the \emph{init} rule, which incorporates Coq's conversion.

It may be at first surprising that there are no
introduction rules for propositional connectives, nor equality for
that matter. However, one of the beauty of the Calculus of Inductive
Construction is that they are, in fact,  defined inductively and
therefore the \emph{decide} rule will handle those; this also
simplifies the syntax of proof-terms.

@@

\subsubsection{Coq-specific code}
Coq terms are accessed through the Coq-Elpi API, and their representation
in \lP takes advantage of native \lP constructs such as lists and binders.
The following is part of the Coq-Elpi API signature of constants that we use:
\begin{lstlisting}[language=lprolog]
kind term   type.                                 % reification of Coq terms
kind gref   type.                                 % reif. of references to global terms
type global gref -> term.                            % coercion to term
type indt   inductive -> gref.                       % reification of inductive types
type indc   constructor -> gref.                     % reification of their constructors 
type app    list term -> term.                       % reification of nary application
type prod   name -> term -> (term -> term) -> term. % reification of dependent product
\end{lstlisting}
%
Note that \lsti{prod} encodes dependent products by taking a name for pretty
printing, a term and a \lP abstraction from terms to terms: i.e., $(x :
B) D$ is encoded by \verb|prod ``x'' B (x\ D x)|; when, in the
implementation of the product-left rule, we apply \lsti{D} to the variable
\lsti{Tm}, we get a new term that can be used to continue backchaining. This
application is obtained via meta-level substitution, in the style of HOAS. In
this sense, our calculus uses \emph{implicit} substitutions, rather than
explicit ones as in the  LJT and PTSC's tradition; this is consistent with proof search in
our application being cut-free, whereas explicit substitutions are linked to cuts.
% Syntax: one can think as if something of type \lsti{term -> term} were a
% \lsti{term} with a hole in it for another \lsti{term}.
The decide rule makes use of the Coq-Elpi primitives
\lsti{coq.safe-dest-app} to obtain the head term of a (possibly
nested) application, and \lsti{coq.env.indt} to access the global
environment of inductive definitions and query for information about them.
The \lsti{decideE} predicate,  has, among others, the role of selecting which
constructor to focus on from the inductive type. The kernel will successively
obtain the type of the selected constructor, and initiate the backchaining
phase. The latter is terminated when the focused atom unifies with the current goal, via
Elpi's primitive \lsti{coq.unify-eq}.

\subsubsection{FPC-specific code}
\label{ssec:fpc}

Each inference rule in Figure~\ref{fig:augmented} is augmented with an
additional premise involving an \emph{expert predicate} and is
annotated with variables $\Xi$, $\Xi'$, $\Xi_1$, $\Xi_2$ that range
over proof certificates.  These expert predicates are used to extract
information from a certificate.  For example, the premise
$\prodE{\XXi}{\XXi_1}{\XXi_2}{t}$ can be understood as a procedure for
extracting from the certificate $\XXi$ a substitution term $t$ and two
continuation certificates $\XXi_1$ and $\XXi_2$.
%
These notions are taken from the general setting of \emph{foundational
  proof certificates}~\cite{chihani17jar}.
%
%%\todo{MM: Talk about randomized FPC? They mentioned afterwards -am}
In our case here, an FPC is a collection of \lP clauses that provide
the remaining details not supplied in Figures~\ref{fig:augmented}
and~\ref{fig:kernel}: that is, the exact set of constructors for the
type of certificates \lsti{cert} as well as the specification of the expert
predicates listed \emph{ibidem}.
%
The top of Figure~\ref{fig:resources} displays two FPCs,
both of which can be used to describe proofs where we bound
the dimension of  a proof.
%
For example, the first FPC dictates that the query \mbox{\lsti{(check (qheight 5) A)}} is
provable in  the kernel using the clauses in Figures~\ref{fig:kernel}
and~\ref{fig:resources} if and only if the height of that proof is 5
or less.
%
Similarly, the second FPC can be used to bound the total number of
instances of the decide rule in a proof.
%
%In particular, the query \mbox{\lsti{(check (qsize 5 H) B)}}
% \begin{lstlisting}
% sigma H\ (check (qsize 5 H) B)
% \end{lstlisting}
%is provable if and only if the total number of unfoldings of that
%proof is 5 or less.

>>>>

\begin{figure}[t]
\centering
  \[
\infer{\XXi\vdash k\ l : A}
      { \mathrm{Ind}[p] \ (\Gamma_I := \Gamma_C) \in E\quad 
       (\mathit{head} \ A) : T \in \Gamma_I\quad  
       k : D \in \Gamma_C\quad 
       \bc{\XXi_1}{} l D A \quad\unfoldExpert{\XXi}{\XXi_1}} 
\]
\vskip -18pt
\[
  \infer{\XXi\vdash  A : s}{E[] \vdash_{CIC} A : s\quad \sortExpert{\XXi}}% The order should be s : A, not A : s ?
  %% we are checking nat : Set, for example, so A : s --AM
  \qquad
  \infer{\bc{\XXi}{}{[]} A A'}{E[]\vdash_{CIC} A =_{\beta\delta\iota\xi}A'\quad\initLExpert{\XXi}}
\]
  \vskip -18pt
\[
  \infer{\bc{\XXi}{}{(t\ ::\ l)}{\depprod{x}{B}{D}}{A}}
        {\XXi_1\vdash t : B\quad
         \bc{\XXi_2}{}l {D[t/x]} A\quad
         \prodE{\XXi}{\XXi_1}{\XXi_2}{t}}
\]
\caption{Specification of a core calculus% of the checking of a proof certificate and the
  % synthesis of a dependently typed $\lambda$-term
  .}
\label{fig:augmented}
\end{figure}

@@

The proof system is parameterized by Coq's global environment $E$,
here a set of constant and inductive definitions; following Coq's
reference manual, inductive definition are denoted by
$\mathrm{Ind}[p] (\Gamma_I := \Gamma_C)$, where $\Gamma_I$ determines
the names and types of the (possibly mutually) inductive type and
$\Gamma_C$ the names and types of its constructors; finally $p$
denotes the number of parameters and plays here no role. We delegate to
Coq's type checking the enforcement of the well-sortedness of
inductive types. The \emph{local context} is empty, since we are only
dealing with types that correspond to Horn clauses, and atomic types
are inductively defined. In fact, we do not have a $\forall$ rule on
the right, although the proof theory would gladly allow it. This means
at the term level that there are no variables in our grammar of
terms. Terms are always applied to a (possibly empty) list of
arguments.  The \emph{decide} rule, % is the equivalent in our
% a \emph{decide} rule in an explicitly focused system
% \cite{liang09tcs}.  
as in the previous proof system for Horn logic,  given an atom,
 selects a clause on which to backchain on: we lookup the
constructors of an inductive definition from the global
environment, one that matches the head symbol of the atom we aim to backchain on, and then
call the latter judgment that will find a correct instantiation, if any. The rules for backchaining include
 the (conflation of the) left introduction rules for
$\forall$ and $\supset$, as well as the \emph{init} rule, which incorporates Coq's conversion.

It may be at first surprising that there are no
introduction rules for propositional connectives, nor equality for
that matter. However, one of the beauties of the Calculus of Inductive
Construction is that they are, in fact, defined inductively and
therefore the \emph{decide} rule will handle those.  Thus, the syntax
of proof-terms is rather simple.

@@

\subsubsection{Coq-specific code}
Coq terms are accessed through the Coq-Elpi API, and their representation
in \lP takes advantage of native \lP constructs such as lists and binders.
The following is part of the Coq-Elpi API signature of constants that we use.
\begin{lstlisting}[language=lprolog]
kind term   type.                                   % reification of Coq terms
kind gref   type.                                   % reification of refs to globals
type global gref -> term.                            % coercion to term
type indt   inductive -> gref.                       % reification of inductive types
type indc   constructor -> gref.                     % reification of their constructors 
type app    list term -> term.                       % reification of nary application
type prod   name -> term -> (term -> term) -> term. % reification of dependent product
\end{lstlisting}
%
Note that \lsti{prod} encodes dependent products by taking a name for pretty
printing, a term and a \lP abstraction from terms to terms: i.e., $(x :
B) D$ is encoded by \verb|prod ``x'' B (x\ D x)|; when, in the
implementation of the product-left rule, we apply \lsti{D} to the variable
\lsti{Tm}, we get a new term that can be used to continue backchaining. This
application is obtained via meta-level substitution, in the style of HOAS. In
this sense, our calculus uses \emph{implicit} substitutions, rather than
explicit ones as in the  LJT and PTSC's tradition; this is consistent with proof search in
our application being cut-free, whereas explicit substitutions are linked to cuts.
% Syntax: one can think as if something of type \lsti{term -> term} were a
% \lsti{term} with a hole in it for another \lsti{term}.
The decide rule makes use of the Coq-Elpi primitives
\lsti{coq.safe-dest-app} to obtain the head term of a (possibly
nested) application, and \lsti{coq.env.indt} to access the global
environment of inductive definitions and query for information about them.
The \lsti{decideE} predicate,  has, among others, the role of selecting which
constructor to focus on from the inductive type. The kernel will successively
obtain the type of the selected constructor, and initiate the backchaining
phase. The latter is terminated when the focused atom unifies with the current goal, via
Elpi's primitive \lsti{coq.unify-eq}.

\newpage
\subsubsection{FPC-specific code}
\label{ssec:fpc}

The \emph{foundational proof certificates} (FPC) framework was
proposed in \cite{chihani17jar} as a flexible approach to specifying a
range of proof structures in first-order classical and intuitionistic
logics.
%
Such specifications are also executable using a simple logic
programming interpreter.
%
As a result of using logic programming, proof certificates in this
framework are allowed to drop details that could be reconstructed during
the checking phase.
%
For example, substitution instances of quantifiers do not need to be
explicitly described within a certificate since unification within the
logic programming checker is often capable of reconstructing such
substitutions.

In this and the next section, we shall only use a much reduced subset
of the FPC framework.
%
In particular, a proof certificate, denoted by the schematic variable
$\XXi$, is a particular term that is threaded throughout a logic
programming interperter.
%
For example, the inference rules in Figure~\ref{fig:augmented} are
augmented with an additional premise called an \emph{expert predicate}
with are responsible for extracting relevant information from a proof
certificate $\XXi$ as well as constructing continuation certificates,
such as, $\XXi_1$ and $\XXi_2$.
%
For example, the premise $\prodE{\XXi}{\XXi_1}{\XXi_2}{t}$ can be
understood as a procedure for extracting from the certificate $\XXi$ a
substitution term $t$ and two continuation certificates $\XXi_1$ and
$\XXi_2$ for the two premises of this rule.
%
If the certificate $\XXi$ does not contain an explicit substitution
term, the expert predicate can simply return a logic variable which would
denote any term that satisfies subsequent unification problems arising
in completing the check of this certificate.

***

%
%
%%\todo{MM: Talk about randomized FPC? They mentioned afterwards -am}
In our case here, an FPC is a collection of \lP clauses that provide
the remaining details not supplied in Figures~\ref{fig:augmented}
and~\ref{fig:kernel}: that is, the exact set of constructors for the
type of certificates \lsti{cert} as well as the specification of the expert
predicates listed \emph{ibidem}.
%
The top of Figure~\ref{fig:resources} displays two FPCs,
both of which can be used to describe proofs where we bound
the dimension of  a proof.
%
For example, the first FPC dictates that the query \mbox{\lsti{(check (qheight 5) A)}} is
provable in  the kernel using the clauses in Figures~\ref{fig:kernel}
and~\ref{fig:resources} if and only if the height of that proof is 5
or less.
%
Similarly, the second FPC can be used to bound the total number of
instances of the decide rule in a proof.
%
%In particular, the query \mbox{\lsti{(check (qsize 5 H) B)}}
% \begin{lstlisting}
% sigma H\ (check (qsize 5 H) B)
% \end{lstlisting}
%is provable if and only if the total number of unfoldings of that
%proof is 5 or less.
